Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-21 doc/thesis/discussion.Rnw
Almooooooost!

%<<setup, include=FALSE>>=
%    source("global.R")
%@

@@

The preferential attachment power \gls{alpha} had a very strong influence on
tree shape in the range of values we considered
(\cref{fig:alphatrees,fig:kpca}). Although the tree kernel was the most
effective classifier for \gls{alpha}, a Sackin's index of tree imbalance
performed nearly as well (\cref{fig:rsquared}). This result was intuitive: high
\gls{alpha} values produce networks with few well-connected ``superspreader''
nodes which are involved in a large number of transmissions, resulting in a
highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
\del{There was no observable bias in the estimates of \gls{alpha} obtained with
\software{netabc}, however the variation in these estimates
was higher for $\alpha < 1$ than for $\alpha \geq 1$} \add{There appeared to be
weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha} $\geq$ 1}
(\cref{fig:abcptm2,tab:abchpd}), which may be partially explained by the
relationship between \gls{alpha} and the power law exponent \gls{gamma}
(\cref{fig:gamma}). \add{Although the degree distributions do not truly follow
a power law for $\gls{alpha} \neq 1$, the fitted exponent still captures the
slope of the degree distribution reasonably well (\cref{fig:powerlaw}), and 
distributions with similar fitted \gls{gamma} values appear qualitatively
similar in other respects.} The \gls{gamma} values fitted to \gls{alpha} = 0
and \gls{alpha} = 0.5 are nearly identical (about 2.28 for \gls{alpha} = 0 and
2.33 for \gls{alpha} = 0.5 with \gls{N} = 5000 and \gls{m} = 2). In other
words, the degree distributions of networks with \gls{alpha} $< 1$ are similar
to each other, which may result in similarity of corresponding transmission
trees as well.

@@

\gls{I} and \gls{N} were both systematically over-estimated by
\software{netabc}, although the bias was more severe for \gls{N} than for
\gls{I}. The number of infected individuals follows a logistic growth curve
under the \gls{SI} model. This kind of growth curve has three qualitative
phases: a slow ramp-up, an exponential growth phase, and a slow final phase
when the susceptible population is almost depleted. The waiting times until the
next transmission, which determine the coalescence times in the tree, are
dependent on the growth phase of the epidemic. Therefore, we hypothesize that
it is the growth phase at the time of sampling which most affects tree shape,
rather than the specific values of \gls{I} or \gls{N}. To investigate this
hypothesis, we simulated transmission trees over networks on a grid of \gls{I}
and \gls{N} values in the region of uniform prior density. We fit logistic
growth curves to the proportion of infected individuals over time, and
calculated the first and second derivatives of these curves (with respect to
time) at the time of transmission tree sampling. These derivatives give us an
indication of the growth rates of the epidemics at the time of sampling. As
shown in \cref{fig:derivs}, there are bands along which both derivatives are
similar which contain the values we tested. These bands span mostly higher
values of \gls{N} and \gls{I} than the true values. Therefore, if \gls{N} and
\gls{I} are free to vary (as is the case in \gls{ABC}), and our hypothesis is
true, both parameters will tend to be overestimated due to being less
identifiable within their own band. However, when \gls{N} is fixed at 5000, the
derivatives vary substantially along the \gls{I}-axis, which explains why
the grid search estimates of \gls{I} were accurate and unbiased
(\cref{fig:gridptI}). \add{We also note the resemblance of the contour surface
of \cref{fig:derivs} to the two-dimensional marginal posterior distributions on
\gls{I} and \gls{N} obtained with simulated data
(\cref{fig:abclow2,fig:abchigh2}). However, while fixing \gls{N} and any other
parameter resulted in much more accurate estimates of \gls{I}, fixing \gls{N}
alone did not (\cref{fig:marginal}). Therefore, there may be more complex
parameter dependence in the \gls{BA} model which we have not considered here.}

\begin{figure}[ht]
  \centering
  %\includegraphics{derivatives.pdf}
  \caption[
    First and second time derivatives of epidemic growth curves at time of
    sampling for various values of $I$ and $N$.
  ]{
    First and second \add{time} derivatives of epidemic growth curves at time
    of sampling for various values of \gls{I} and \gls{N}. Networks were
    simulated under the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and
    \gls{N} varied along the values shown on the $x$-axis. Transmission trees
    were sampled at the time when \gls{I} nodes were infected ($y$-axis).
    Logistic growth curves were fit to epidemic trajectories derived from the
    transmission trees, and their first and second derivatives were calculated
    at the time of sampling. Contours show first derivatives, while colours
    indicate second derivatives. Values of \gls{I} and \gls{N} used in
    simulation experiments with \gls{ABC} are indicated by diamonds.
  }
  \label{fig:derivs}
\end{figure}

@@

\end{document}

@@

\add{For all datasets we examined, the posterior mean estimates for \gls{alpha}
were sub-linear, ranging from 
    \Sexpr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
to 
    \Sexpr{d[m_min == 1 & parameter == "alpha", round(max(mean), 2)]}.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. They found \gls{alpha} values
ranging from 0.26 to 0.62, depending on the gender and time period considered. 
Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
this range under both priors, as were the estimates for the
\textcite{wang2015targeting} data and the BC data when $\gls{m} = 1$ was
disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.}

Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
populations whose primary risk factor for \gls{HIV} infection was heterosexual
contact. \citeauthor{de2007preferential} explicitly excluded reported
homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
heterosexual contact is the primary mode of transmission in Botswana where the
study was done. In the first of the two papers describing the Botswana
study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
was gender-biased, being composed of approximately 75\% women. Our estimate of
\gls{alpha} for these data was 
    X%\S%expr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
        or 
    X%\S%expr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
depending on the prior on \gls{m}; \textcite{de2007preferential} estimated
0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership networks
respectively for the female portion of their sample.

For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
populations had a higher estimated \acrlong{PA} power than the other datasets
(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
in Brooklyn, USA, collected between 1991 and
1993~\autocite{friedman2006social}. They found that the the \gls{IDU} network 
resembled a \gls{BA} network much more closely than other social and sexual 
networks, and offered sociological explanations for the apparent \acrlong{PA}
dynamics in this population. Importantly, from a public health perspective,
the authors asserted that the removal of \emph{random} individuals from
\gls{IDU} networks may have the paradoxical effect of increasing the network's
epidemic susceptibility. When low-degree nodes are removed, as would occur
during a police crackdown, their network neighbours may turn to well-known
community members for advice or supplies, thus increasing the connectivity of
these high-degree nodes.

@@

%One somewhat surprising result was the difference between parameter estimates 
%for the \textcite{li2015hiv} and \textcite{wang2015targeting} datasets. Both
%groups studied cohorts of acutely infected \gls{MSM} in major Chinese cities
%%(Shanghai and Beijing respectively); yet, the \citeauthor{li2015hiv} data was 
%estimated to have a lower \acrlong{PA} power and larger infected population
%than the \textcite{wang2015targeting} data\ldots

\subsubsection{Power law exponent}

% for discussion
In order to compare our results to existing literature on networks and
distributions of partner counts, we have reported estimated values for the
power law exponent \gls{gamma} of the real data sets we evaluated. However, the
posterior means for \gls{alpha} for all six datasets were less than one; the
degree distributions in this parameter range are stretched exponential, not
power law~\autocite{krapivsky2000connectivity}. As we show in
\cref{fig:powerlaw}, the power law fit does capture the slope of the degree
distribution fairly well, but the results should still be interpreted
cautiously. \textcite{krapivsky2000connectivity} showed that the power law
distribution can be maintained, with \gls{gamma} tuned to any desired value, by
a straightforward modification of the \gls{BA} model. The authors define the
``connection kernel'' $A_k$ the probability of a new connection to a node of
degree $k$, up to a normalizing constant. In the \gls{BA} model as we have
presented it here, $A_k = k^{\alpha} + 1$. Taking $A_k$ as any asymptotically
linear function will result in a power law distribution, with the exponent
\gls{gamma} determined by the properties of $A_k$. Implementing such a model
would be straightforward and seems a natural next step toward improving the
realism of the \gls{BA} model.

\cref{tab:litgamma} shows some estimated \gls{gamma} values reported in the
network literature.  The values mainly fall between 2 and 4, although
substantial variation within this range exists for networks with apparently
similar demographics. Some of these differences are likely due to the differing
time intervals considered in each study -- some investigate the distribution of
lifetime partners, while others consider only recent partnerships, such as in
the past year. TODO

\begin{table}[ht]
    \centering
    \input{\tablepath/lit-gamma}
    \caption[
        Values of power law exponent \gls{gamma} previously reported in the
        literature.
    ]{
        Values of power law exponent \gls{gamma} previously reported in the
        literature. In the risk group column, ``mixed sexual'' indicates
        participants were not selected on the basis of sexual orientation;
        these values can be assumed to reflect primarily heterosexual dynamics.
        \textcite{brown2011transmission} studied a phylogenetically estimated
        transmission network, rather than a contact network.
        The values presented by \textcite{jones2003assessment} were in the
        context of arguing that the power law distribution is not suitable for
        use with network data.
    }
    \label{tab:kernelexpt}
\end{table}

%For both priors, the estimated prevalence was extremely high, in fact higher
%than the estimated HIV prevalence in the sampled region. The authors indicated
%that the source of the samples was a town in close proximity to the country's
%capital city, and suggested that there may have been a high degree of migration
%and partner interchange between the two locations. It is possible that the
%contact network underlying the subtree we investigated includes a much larger
%group based in the capital city, which would explain the high estimate of $I$.
%There is no clear explanation for the discrepancy between the two priors for
%the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
%cluster and therefore was a good candidate for the BA model. However, nearly
%all the posterior density was assigned to $m = 1$ when this value was allowed,
%indicating that the network was more likely to have an acyclic tree structure.

\subsubsection{HIV prevalence}

\add{The true \gls{HIV} prevalence in a population can be
difficult to estimate for several reasons. \gls{HIV}-infected individuals may
be asymptomatic for months or years, possibly delaying their awareness of their
status. In many contexts, the risk factors for acquisition of \gls{HIV} are
illegal or stigmatized, which may represent a barrier to testing, treatment,
and/or disclosure of status. Several of the studies whose data we analysed
report the number of new infections over the study period. We expect these
numbers to be lower, perhaps substantially, than the true prevalence. We also
note that our simulation study showed that the estimates of \gls{I} obtained
with \software{netabc} were upwardly biased, the severity increasing with the
true value of \gls{I}. In addition, our initial exploratory analysis showed
that the identifiability of \gls{I} decreases with the number of sampled tips 
(\cref{fig:Icrossv,fig:Ikpca}). In real world studies, the proportion of
infected individuals sampled is usually low, which may impede our ability to
estimate \gls{I}. In the following, we discuss any information available about
the \gls{HIV} prevalence in the analysed study populations, and relate this to
our estimates. However, due to the aforementioned challenges, this information
cannot be taken as absolute ``ground truth''. Discrepancies between our results
and the available data may indicate that the estimates produced by
\software{netabc} are inaccurate, but they may also indicate that the available
epidemiological data was not sufficient to estimate the true prevalence. Since
the estimates of the prevalence \gls{I} were largely robust to the choice of
prior on \gls{m} (\cref{fig:abchpd,fig:abchpdm2}), we discuss here only the
results obtained with the prior $\gls{m} \sim \DiscreteUniform(1, 5)$.}

\add{The prevalence estimates for two \gls{IDU} epidemics in Eastern Europe,
IDU/Estonia and IDU/Romania, were similar 
    (X and X), 
although the confidence interval around the \textcite{niculescu2015recent} data
was wider. \textcite{niculescu2015recent} reported that 494 new \gls{HIV}
infections were diagnosed among the studied population (\glspl{IDU} in
Bucharest) during the study period, and this value was contained in our 95\%
\gls{HPD} 
    (X-X). 
On the other hand, \textcite{zetterberg2004two} collected samples from several
locations throughout Estonia; using their reported incidence estimates, there
were approximately 2000 new infections in the country during the study period.
This value was outside our 95\% \gls{HPD} 
    (X-X). 
The authors indicated that their sample was ``convenience based''; it is
possible that other, harder-to-reach individuals had no contacts with those
sampled, reducing the apparent network size. In contrast, the IDU/Canada data
was collected primarily from \gls{IDU} in Vancouver's downtown east side, a
population that has been intensively targeted by public health interventions.
Combined with the routine clinical genotyping performed in British Columbia for
all new \gls{HIV} infections, it is not unreasonable to suppose that the
sampled individuals (399) comprise the majority of \gls{HIV}-positive network
members. TODO}

\add{Surprisingly, the prevalence estimates for two \gls{MSM} populations major
Chinese cities (MSM/Shanghai and MSM/Beijing) were quite different
    (X and X).
The individuals sampled for the MSM/Beijing study were recruited by following a
prospective cohort of size 2000. 179 new infections were identified in the
cohort during the study period, which is much lower than the estimated
prevalence obtained with \software{netabc}
    (X).
However, the authors did not claim to have sampled the entire sexual network.
In a nationwide survey, \textcite{wu2013hiv} estimated that there were 24,198
\gls{MSM} living in Beijing, of whom 5.7\%, or 1379, were \gls{HIV}-infected,
which is within the 95\% confidence interval of our estimate of \gls{I}
    (X-X).
The same survey estimated the size of the \gls{MSM} population in Shanghai
14,511, with an \gls{HIV} prevalence of 6.8\% or approximately 987 individuals.
This was within the 95\% confidence interval of the estimate obtained with
\software{netabc} for the MSM/Shanghai data
    (X-X). 
It is worth noting that the authors of the IDU/Shanghai
study~\autocite{li2015hiv} claimed that there were 80,000 \gls{MSM} living in
Shanghai, which is a significantly different estimate than that obtained by
\textcite{wu2013hiv}. The authors of the MSM/Taiwan
study~\autocite{kao2011surveillance} stated that there were 18,378
\gls{HIV}-infected individuals in Taiwan, of which approximately 44\% (8086
individuals) were \gls{MSM}. This value is well outside of our 95\% \gls{HPD} 
    (X-X),
suggesting that the entire \gls{MSM} population does not comprise a single
network. In fact, our estimate of \gls{I} for these data
    (X)
was closer to the number of \gls{MSM} enrolled in the study, 301. The study
focused exclusively on newly diagnosed individuals; perhaps chronically
infected \gls{MSM} were no longer part of the same network. On the other hand,
our estimated prevalence for the MSM/USA data
    (X)
was clearly an underestimate. The study enrolled almost 648 \gls{HIV} infected 
individuals, which should have been a lower bound on \gls{I}, although the
value was contained in the 95\% \gls{HPD} interval
    (X-X).}

>>>>

<<setup, include=FALSE>>=
    source("global.R")
@

@@

The preferential attachment power \gls{alpha} had a very strong influence on
tree shape in the range of values we considered
(\cref{fig:alphatrees,fig:kpca}). Although the tree kernel was the most
effective classifier for \gls{alpha}, a Sackin's index of tree imbalance
performed nearly as well (\cref{fig:rsquared}). This result was intuitive: high
\gls{alpha} values produce networks with few well-connected ``superspreader''
nodes which are involved in a large number of transmissions, resulting in a
highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
\del{There was no observable bias in the estimates of \gls{alpha} obtained with
\software{netabc}, however the variation in these estimates
was higher for $\alpha < 1$ than for $\alpha \geq 1$} \add{There appeared to be
weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha} $\geq$ 1}
(\cref{fig:abcptm2,tab:abchpd}), which may be partially explained by the
relationship between \gls{alpha} and the power law exponent \gls{gamma}
(\cref{fig:gamma}). \add{Although the degree distributions do not truly follow
a power law for $\gls{alpha} \neq 1$, the fitted exponent still captures the
slope of the degree distribution reasonably well (\cref{fig:powerlaw}), and 
distributions with similar fitted \gls{gamma} values appear qualitatively
similar in other respects.} The \gls{gamma} values fitted to $\gls{alpha} = 0$
and $\gls{alpha} = 0.5$ are nearly identical (about $2.28 for \gls{alpha} = 0$
and $2.33 for \gls{alpha} = 0.5$ with $\gls{N} = 5000$ and $\gls{m} = 2$). In
other words, the degree distributions of networks with \gls{alpha} $< 1$ are
similar to each other, which may result in similarity of corresponding
transmission trees as well.

@@

\gls{I} and \gls{N} were both systematically over-estimated by
\software{netabc}, although the bias was more severe for \gls{N} than for
\gls{I}. The number of infected individuals follows a logistic growth curve
under the \gls{SI} model. This kind of growth curve has three qualitative
phases: a slow ramp-up, an exponential growth phase, and a slow final phase
when the susceptible population is almost depleted. The waiting times until the
next transmission, which determine the coalescence times in the tree, are
dependent on the growth phase of the epidemic. Therefore, we hypothesize that
it is the growth phase at the time of sampling which most affects tree shape,
rather than the specific values of \gls{I} or \gls{N}. To investigate this
hypothesis, we simulated transmission trees over networks on a grid of \gls{I}
and \gls{N} values in the region of uniform prior density. We fit logistic
growth curves to the proportion of infected individuals over time, and
calculated the first and second derivatives of these curves (with respect to
time) at the time of transmission tree sampling. These derivatives give us an
indication of the growth rates of the epidemics at the time of sampling. As
shown in \cref{fig:derivs}, there are bands along which both derivatives are
similar which contain the values we tested. These bands span mostly higher
values of \gls{N} and \gls{I} than the true values. Therefore, if \gls{N} and
\gls{I} are free to vary (as is the case in \gls{ABC}), and our hypothesis is
true, both parameters will tend to be overestimated due to being less
identifiable within their own band. However, when \gls{N} is fixed at 5000, the
derivatives vary substantially along the \gls{I}-axis, which explains why
the grid search estimates of \gls{I} were accurate and unbiased
(\cref{fig:gridest,fig:gridptI}). \add{We also note the resemblance of the
contour surface of \cref{fig:derivs} to the two-dimensional marginal posterior
distributions on \gls{I} and \gls{N} obtained with simulated data
(\cref{fig:abclow2,fig:abchigh2}). In addition, both the accuracy and precision
of the \gls{ABC} estimates of \gls{I} improved substantially when \gls{N} was
marginalized out (\cref{fig:marginal}).}

\begin{figure}[ht]
  \centering
  \includegraphics{derivatives.pdf}
  \caption[
    First and second time derivatives of epidemic growth curves at time of
    sampling for various values of $I$ and $N$.
  ]{
    First and second \add{time} derivatives of epidemic growth curves at time
    of sampling for various values of \gls{I} and \gls{N}. Networks were
    simulated under the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and
    \gls{N} varied along the values shown on the $x$-axis. Transmission trees
    were sampled at the time when \gls{I} nodes were infected ($y$-axis).
    Logistic growth curves were fit to epidemic trajectories derived from the
    transmission trees, and their first and second derivatives were calculated
    at the time of sampling. Contours show first derivatives, while colours
    indicate second derivatives. Values of \gls{I} and \gls{N} used in
    simulation experiments with \gls{ABC} are indicated by diamonds.
  }
  \label{fig:derivs}
\end{figure}

@@

@@

\add{For all datasets we examined, the posterior mean estimates for \gls{alpha}
were sub-linear, ranging from 
    \Sexpr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
to 
    \Sexpr{d[m_min == 1 & parameter == "alpha", round(max(mean), 2)]}.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. When used to analyse
population-level longitudinal partner count data, they found \gls{alpha} values
ranging from 0.26 to 0.62 depending on the gender and time period considered.
Our estimates of \gls{alpha} for the IDU/Romania was above this range under
both priors, as were the estimates for the MSM/Beijing data and the BC data
when $\gls{m} = 1$ was disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.}

\add{Both \textcite{de2007preferential} and the HET/Botswana data studied
populations whose primary risk factor for \gls{HIV} infection was heterosexual
contact. \citeauthor{de2007preferential} explicitly excluded reported
homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
heterosexual contact is the primary mode of transmission in Botswana where the
study was done. In the first of the two papers describing the Botswana
study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
was gender-biased, being composed of approximately 75\% women. Our estimate of
\gls{alpha} for these data was 
    \Sexpr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
        or 
    \Sexpr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
depending on the prior on \gls{m}. Similarly, \textcite{de2007preferential}
estimated 0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership
networks respectively for the female portion of their sample.}

\add{For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
populations had a higher estimated \acrlong{PA} power than the other datasets
(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
in Brooklyn, USA, collected between 1991 and
1993~\autocite{friedman2006social}. They found that the \gls{IDU} network
resembled a \gls{BA} network much more closely than other social and sexual 
networks, and offered sociological explanations for the apparent \acrlong{PA}
dynamics in this population. Importantly, from a public health perspective,
the authors asserted that the removal of \emph{random} individuals from
\gls{IDU} networks may have the paradoxical effect of increasing the network's
epidemic susceptibility. When low-degree nodes are removed, as would occur
during a police crackdown, their network neighbours may turn to well-known
community members for advice or supplies, thus increasing the connectivity of
these high-degree nodes.}

@@

\subsubsection{Power law exponent is within reported range, but hard to interpret}

\add{In order to compare our results to existing literature on networks and
distributions of partner counts, we have reported estimated values for the
power law exponent \gls{gamma} of the real data sets we evaluated. However, the
posterior means for \gls{alpha} for all six datasets were less than one; the
degree distributions in this parameter range are stretched exponential, not
power law~\autocite{krapivsky2000connectivity}. As we show in
\cref{fig:powerlaw}, the power law fit does capture the slope of the degree
distribution fairly well, but the results should still be interpreted
cautiously. \textcite{krapivsky2000connectivity} showed that the power law
distribution can be maintained, with \gls{gamma} tuned to any desired value, by
a straightforward modification of the \gls{BA} model. The authors define the
``connection kernel'' $A_k$ the probability of a new connection to a node of
degree $k$, up to a normalizing constant. In the \gls{BA} model as we have
presented it here, $A_k = k^{\alpha} + 1$. Taking $A_k$ as any asymptotically
linear function will result in a power law distribution, with the exponent
\gls{gamma} determined by the properties of $A_k$. Implementing such a model
would be straightforward and seems a natural next step toward improving the
realism of the \gls{BA} model.}

\add{\Cref{tab:litgamma} shows some estimated \gls{gamma} values reported in the
network literature.  Our estimates are most similar to those of
\textcite{liljeros2001web, rothenberg2007large}, but interpretation of the
exact values is challenging. Estimates from the literature mainly fall between
2 and 4, although substantial variation within this range exists, even for
networks with apparently similar demographics. For example,
\textcite{schneeberger2004scale} estimate higher \gls{gamma} values for
heterosexual females than heterosexual males, while
\textcite{clemenccon2015statistical} find the opposite. Some of these
differences are likely due to the differing time intervals considered in each
study -- some investigate the distribution of lifetime partners, while others
consider only recent partnerships, such as in the past year. However, the
effect of the time period on \gls{gamma} is not readily obvious from the
available estimates.}

\begin{table}[ht]
    \centering
    \input{\tablepath/lit-gamma}
    \caption[
        Values of power law exponent \gls{gamma} previously reported in the
        literature.
    ]{
        Values of power law exponent \gls{gamma} previously reported in the
        literature. In the risk group column, ``mixed sexual'' indicates
        participants were not selected on the basis of sexual orientation;
        these values likely reflect primarily heterosexual dynamics.
        \textcite{brown2011transmission} studied a phylogenetically estimated
        transmission network, rather than a contact network.
        The values presented by \textcite{jones2003assessment} were in the
        context of arguing that the power law distribution is not suitable for
        use with network data.
    }
    \label{tab:litgamma}
\end{table}

\subsubsection{HIV prevalence estimates are often discordant with epidemiological information}

\add{The true \gls{HIV} prevalence in a population can be
difficult to estimate for several reasons. \gls{HIV}-infected individuals may
be asymptomatic for months or years, possibly delaying their awareness of their
status. In many contexts, the risk factors for acquisition of \gls{HIV} are
illegal or stigmatized, which may represent a barrier to testing, treatment,
and/or disclosure of status. Several of the studies whose data we analysed
report the number of new infections over the study period. We expect these
numbers to be lower, perhaps substantially, than the true prevalence. We also
note that our simulation study showed that the estimates of \gls{I} obtained
with \software{netabc} were upwardly biased, the severity increasing with the
true value of \gls{I}. In addition, our initial exploratory analysis showed
that the identifiability of \gls{I} decreases with the number of sampled tips 
(\cref{fig:Icrossv,fig:Ikpca}). In real world studies, the proportion of
infected individuals sampled is usually low, which may impede our ability to
estimate \gls{I}. In the following, we discuss any information available about
the \gls{HIV} prevalence in the analysed study populations, and relate this to
our estimates. However, due to the aforementioned challenges, this information
cannot be taken as absolute ``ground truth''. Discrepancies between our results
and the available data may indicate that the estimates produced by
\software{netabc} are inaccurate, but they may also indicate that the available
epidemiological data was not sufficient to estimate the true prevalence. Since
the estimates of \gls{I} were largely robust to the choice of prior on \gls{m}
(\cref{fig:abchpd,fig:abchpdm2}), we discuss here only the results obtained
with the prior $\gls{m} \sim \DiscreteUniform(1, 5)$.}

\add{The prevalence estimates for two \gls{IDU} epidemics in Eastern Europe,
IDU/Romania and IDU/Estonia, were 
    \Sexpr{d1[dataset == "niculescu2015" & parameter == "I", round(mean)]} and 
    \Sexpr{d1[dataset == "zetterberg2004" & parameter == "I", round(mean)]}.
The authors of the IDU/Romania study~\autocite{niculescu2015recent} reported
that 494 new \gls{HIV} infections were diagnosed among the studied population
(\glspl{IDU} in Bucharest) during the study period, and this value was
contained in our 95\% \gls{HPD} interval
    (\Sexpr{d1[dataset == "niculescu2015" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "niculescu2015" & parameter == "I", round(hpd.max.95)]}).
On the other hand, the IDU/Estonia data were collected from several locations
in the country. Using their reported incidence estimates, there were
approximately 2000 new infections in the country during the study period, a
value outside our 95\% \gls{HPD} interval
    (\Sexpr{d1[dataset == "zetterberg2004" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "zetterberg2004" & parameter == "I", round(hpd.max.95)]}).
The authors indicated that their sample was ``convenience based''; it is
possible that other, harder-to-reach individuals had no contacts with those
sampled, reducing the apparent network size. In contrast, the IDU/Canada data
was collected primarily from \gls{IDU} in Vancouver's downtown east side, a
population that has been intensively targeted by public health interventions.
Combined with the routine clinical genotyping performed in British Columbia for
all new \gls{HIV} infections, it is not unreasonable to suppose that the
sampled individuals (399) comprise the majority of \gls{HIV}-positive network
members. This value was the lower bound of our estimated 95\% \gls{HPD}
interval
    (\Sexpr{d1[dataset == "bctree" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "bctree" & parameter == "I", round(hpd.max.95)]}),
but it seems likely that the posterior mean of
    \Sexpr{d1[dataset == "bctree" & parameter == "I", round(mean)]}
is an overestimate.}

\add{The prevalence estimates for two \gls{MSM} populations in major Chinese
cities (MSM/Beijing and MSM/Shanghai) were 
    \Sexpr{d1[dataset == "wang2015" & parameter == "I", round(mean)]} and
    \Sexpr{d1[dataset == "li2015" & parameter == "I", round(mean)]}.
The individuals sampled for the MSM/Beijing study were recruited by following a
prospective cohort of size 2000. 179 new infections were identified in the
cohort during the study period, which is much lower than the estimated
prevalence obtained with \software{netabc}. However, the authors did not claim
to have sampled the entire sexual network. In a nationwide survey,
\textcite{wu2013hiv} estimated that there were 24,198 \gls{MSM} living in
Beijing, of whom 5.7\%, or 1379, were \gls{HIV}-infected, which is close to the
upper limit of the 95\% confidence interval of our estimate of \gls{I}
    (\Sexpr{d1[dataset == "wang2015" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "wang2015" & parameter == "I", round(hpd.max.95)]}).
The same survey estimated the size of the \gls{MSM} population in Shanghai
14,511, with an \gls{HIV} prevalence of 6.8\% or approximately 987 individuals.
This was within the 95\% confidence interval of the estimate obtained with
\software{netabc} for the MSM/Shanghai data
    (\Sexpr{d1[dataset == "li2015" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "li2015" & parameter == "I", round(hpd.max.95)]}).
It is worth noting that the authors of the IDU/Shanghai
study~\autocite{li2015hiv} claimed that there were 80,000 \gls{MSM} living in
Shanghai, which is a significantly different estimate than that obtained by
\textcite{wu2013hiv}. The authors of the MSM/Taiwan
study~\autocite{kao2011surveillance} stated that there were 18,378
\gls{HIV}-infected individuals in Taiwan, of which approximately 44\% (8086
individuals) were \gls{MSM}. This value is well outside of our 95\% \gls{HPD} 
    (\Sexpr{d1[dataset == "kao2011" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "kao2011" & parameter == "I", round(hpd.max.95)]}),
suggesting that the entire \gls{MSM} population does not comprise a single
network. In fact, our estimate of \gls{I} for these data
    (\Sexpr{d1[dataset == "kao2011" & parameter == "I", round(mean)]}),
was closer to the number of \gls{MSM} enrolled in the study (301) than to the
epidemiologically estimated prevalence. The study focused exclusively on newly
diagnosed individuals. On the other hand, our estimated prevalence for the
MSM/USA data
    (\Sexpr{d1[dataset == "little2014" & parameter == "I", round(mean)]})
was clearly an underestimate. The study enrolled almost 648 \gls{HIV} infected 
individuals, which should have been a lower bound on \gls{I}, although the
value was contained in the 95\% \gls{HPD} interval
    (\Sexpr{d1[dataset == "little2014" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "little2014" & parameter == "I", round(hpd.max.95)]}).}

\add{The HET/Uganda, HET/Malawi, and HET/Botswana data all originated from
populations in sub-Saharan Africa, where the dominant mode of transmission is
believed to be heterosexual. The HET/Uganda study population included
1434 seropositive individuals. Since the majority of inhabitants of the study
district were enrolled in the study, it is likely that this number represents
the majority of \gls{HIV} infected persons. Our estimate was slightly lower
    (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(mean)]}),
but the value 1434 was contained ith the 95\% \gls{HPD} interval
    (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(hpd.max.95)]}).
The HET/Malawi data represents possibly the oldest population-level \gls{HIV}
sequencing effort, carried out on samples collected for other purposes between
1981 and 1989. Every individual in the study district was sampled; among these,
a total of 200 were \gls{HIV}-positive. Although the HET/Malawi data had the
lowest estimated \gls{I} value among the heterosexual datasets, 
the posterior mean
    (\Sexpr{d1[dataset == "mccormack2002" & parameter == "I", round(mean)]}),
was much higher than this value. Due to the thorough population-level
screening, it is most likely that the posterior mean represents an
overestimate, especially considering the observed upward bias of \gls{ABC}
estimates of \gls{I} observed on simulated data (\cref{fig:abcptm2}).
It is also possible that the true sexual network extended outside the study's
geographic area to include a larger number of infected individuals. The authors
of the HET/Botswana data~\autocite{novitsky2013phylogenetic,
novitsky2014impact} estimated that there were 1731 \gls{HIV}-positive
individuals in the study area. \gls{HIV} sequences were obtained from
approximately 70\% of these individuals. The estimated prevalence we obtained
was much higher 
    (\Sexpr{d1[dataset == "novitsky2014" & parameter == "I", round(mean)]}),
with a 95\% \gls{HPD} interval spanning nearly the entire prior region. The
authors noted that the town where the study took place was located proximally
to the country's capital city, and suggested that frequent travel between the
two locations may have facilitated linking of their sexual networks. Thus, the
high estimate of \gls{I} we obtained for these data may include a larger
network component based in the capital city.}
